The Effects of weather on maize yield: New evidence from Kenya

### 2019-01-13 writing/draft4/draft20.tex
MAPE asal non asal

According to the magnitude of the standardised coefficients and F-values in the first two columns of table~\ref{MainEst}, the most important variable is average seasonal temperature which appears to be much more important than the remaining predictors. The second most important variable is total seasonal precipitation. The value of the F-statistic corresponding to temperature is $12.010$ while the F-statistic of the linear term of cumulative precipitation is equal to $5.358$ and the F-statistic of its quadratic term is $4.263$. Similarly, according to the Satterthwaite's p-values, average temperature is the most important predictor as the corresponding Satterthwaite's p-value is equal to $6\times10^{-4}$ which is much lower than the p-values of the other explanatory variables.

We will now discuss and compare the results of the models based on the subsample of the ASAL counties (the third and fourth columns of table~\ref{MainEst}) and the model based on the subsample of the non-ASAL counties (the fifth and sixth columns of table~\ref{MainEst}). Interestingly, different predictors were found significant for each of the two subsamples. Regarding the subsample of the ASAL counties, the only significant fixed effects are average seasonal temperature and length and number of dry spells. According to our expectation, all three variables have a negative sign. On the other hand, the variables which were found significant for the subsample of the non-ASAL counties include seasonal precipitation (positive) and its square (negative) and CV of temperature (positive). According to the F-statistics, the most important explanatory variables for the ASAL counties are number of dry spells and length of the longest dry spell with values of the F-statistics equal to $8.032$ and $6.841$, respectively. The F-statistic of the average temperature is $5.409$ and the F-statistics of the remaining predictors are lower than zero. However, according to the magnitude of the standardised coefficients, the average temperature would be the most important as the coefficient is equal to $-0.214$ while it is equal to $-0.156$ and $-0.181$ for the number of dry spells and the length of maximum dry spell, respectively. With regard to the subsample of the non-ASAL counties, the most important explanatory variables based on the F-statistic are precipitation and its square. The F-statistic of the squared values of precipitation is $23.384$ and its value for the linear term of precipitation is $19.140$, while the variable with the third highest F-statistic is CV of temperature with the F-statistic equal to $5.372$, hence, the difference in importance of the precipitation and CV of temperature is relatively large. The standardised coefficient of squared precipitation is equal to $-0.127$, the coefficient of the linear term of precipitation is $0.244$ and the standardised coefficient of CV of temperature is $0.058$. In terms of the magnitude of the coefficients, the average temperature is also relatively important as the value of the coefficient is $-0.128$. However, average temperature is insignificant according to the Satterthwaite's p-value and its F-statistic is also relatively small, in particular $1.528$.

To sum up, for the ASAL counties temperature is much more important than for the non-ASAL counties where temperature is not significant. On the other hand, precipitation is not significant for the ASAL counties while it is strongly significant for the non-ASAL counties. This is illustrated in figure~\ref{MarEff1}(c)-(f) in this section. The plot of the marginal effect of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(c) in this section) while it is hill-shaped and steep for the non-ASAL counties (figure \ref{MarEff1}(e) in this section) similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f) in this section, the marginal effect of temperature is linear and decreasing for the case of ASAL and non-ASAL counties. Marginal effects of the remaining predictors are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted variant is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the statistic in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects components and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In this case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties.

@@

To check for presence of multicollinearity we calculated the variance inflation factors (VIF) of the explanatory variables. With only two exceptions, the VIF were below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF was $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and $2.182$ in the model for the subsample of the non-ASAL counties. These values are still well below the threshold of problematic multicollinearity. The other exception was the seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably as a result of the presence of the squared of precipitation. The VIF is below $2$ in the other two models and even the value $2.407$ is below the threshold of the problematic multicollinearity. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

\begin{itemize}
\color{blue}
\item Verbal description and interpretation of the results. Discussing goodness of fit using various criteria such as AIC or alternatives to $R^2$.

\item relative importance of the individual variables and its measures
%\item Possibly estimate simple regressions with the same explanatory variables as the preferred mixed effects specification. This would give us an approximation of $R^2$ which could be used to compare our models.

>>>>

According to the magnitude of the standardised regression coefficients and F-values in the first two columns of table~\ref{MainEst}, the most important variable is average seasonal temperature which appears to be much more important than the remaining predictors. The second most important variable is total seasonal precipitation. The value of the F-statistic corresponding to temperature is $12.010$ while the F-statistic of the linear term of cumulative precipitation is equal to $5.358$ and the F-statistic of its quadratic term is $4.263$. Similarly, according to the Satterthwaite's p-values, average temperature is the most important predictor as the corresponding Satterthwaite's p-value is equal to $6\times10^{-4}$ which is much lower than the p-values of the other explanatory variables.

We will now discuss and compare the results of the models based on the subsample of the ASAL counties (the third and fourth columns of table~\ref{MainEst}) and the model based on the subsample of the non-ASAL counties (the fifth and sixth columns of table~\ref{MainEst}). Interestingly, different predictors were found significant for each of the two subsamples. Regarding the subsample of the ASAL counties, the only significant fixed effects are average seasonal temperature and length and number of dry spells. According to our expectation, all three variables have a negative sign. On the other hand, the variables which were found significant for the subsample of the non-ASAL counties are seasonal precipitation (positive) and its square (negative) and CV of temperature (positive). According to the F-statistics, the most important explanatory variables for the ASAL counties are number of dry spells and length of the longest dry spell with values of the F-statistics equal to $8.032$ and $6.841$, respectively. The F-statistic of the average temperature is $5.409$ and the F-statistics of the remaining predictors are lower than one. However, according to the magnitude of the standardised regression coefficients, the average temperature is the most important as its coefficient is equal to $-0.214$ while it is equal to $-0.156$ and $-0.181$ for the number of dry spells and the length of maximum dry spell, respectively. With regard to the subsample of the non-ASAL counties, the most important explanatory variables based on the F-statistic are precipitation and its square. The F-statistic of the squared values of precipitation is $23.384$ and its value for the linear term of precipitation is $19.140$, while the variable with the third highest value of the F-statistic is CV of temperature with the F-statistic equal to $5.372$, hence, the difference in importance of the precipitation and CV of temperature is relatively large. The standardised regression coefficient of squared precipitation is equal to $-0.127$, the coefficient of the linear term of precipitation is $0.244$ and the standardised regression coefficient of CV of temperature is $0.058$. In terms of the magnitude of the coefficients, the average temperature is also relatively important as the value of its coefficient is $-0.128$. However, average temperature is insignificant according to the Satterthwaite's p-value and the value of its F-statistic is also relatively low, in particular $1.528$.

To sum up, temperature is much more important for the ASAL counties than for the non-ASAL counties where temperature is isignificant. On the other hand, precipitation is not significant for the ASAL counties while it is strongly significant for the non-ASAL counties. This is illustrated in figures~\ref{MarEff1}(c)-(f) in this section. The plot of the marginal effect of precipitation is almost flat for the ASAL counties (figure \ref{MarEff1}(c) in this section) while it is hill-shaped and steep for the non-ASAL counties (figure \ref{MarEff1}(e) in this section) similarly as in the case of the whole sample. As one can see in figures~\ref{MarEff1}(d) and~\ref{MarEff1}(f) in this section, the marginal effect of temperature is linear and decreasing for the both subsamples . Marginal effects of the remaining predictors are depicted in figures~\ref{MarEff2}(c)-(f) and~\ref{MarEff3}(c)-(f) in the appendix.

@@

The AIC is an important criterion for comparison of competing models, however it does not provide any information about the absolute goodness of fit. In order to assess the absolute goodness of fit of a model, a different measure is needed. For a simple linear regression, the coeffiecient of determination $R^2$ or its adjusted version is typically used for this purpose. Nevertheless, obtaining the $R^2$ or the adjusted $R^2$ is not straightforward for linear mixed models or for models estimated by the maximum likelihood in general. In this study, we used the coefficient of determination of generalized linear mixed-effects models introduced by \cite{Nakagawa2013}. For each of our samples (all counties, ASAL and non-ASAL), we present two versions of the statistic in table~\ref{R2}. These include marginal $R^2$ which can be interpreted as the variance explained by the fixed effects components and the conditional $R^2$ which can be interpreted as the variance explained by both fixed effects and random effects. In our case, the only random effects are the intercepts, hence the difference between the two variants is the variance explained by the random intercepts. As is apparent from table~\ref{R2}, the fixed effects explain about $30\%$ of the variance in case of the model for all counties,  about $13\%$ of the variance in the model based on the ASAL counties and about $15\%$ of the variance in the model based on the non-ASAL counties.

@@

According to a standard practise, we calculated the variance inflation factors (VIF) of the explanatory variables to test for presence of multicollinearity. With only two exceptions, the VIF is below $2$ for all predictors in our final models. For CV of precipitation, the value of VIF is $2.172$ in the model of all counties, $2.147$ in the model for the subsample of the ASAL counties and it is equal to $2.182$ in the model based on the subsample of the non-ASAL counties. These values are well below the threshold of problematic multicollinearity. The other exception is the seasonal precipitation as its VIF is $2.407$ in the model for the non-ASAL counties, probably due to presence of the squared term of precipitation. However, $2.407$ is below the threshold of the problematic multicollinearity. The VIF of squared precipitation is below $2$ in the other two models. Therefore, we can conclude that the models do not suffer from presence of multicollinearity.

%\item Possibly estimate simple regressions with the same explanatory variables as the preferred mixed effects specification. This would give us an approximation of $R^2$ which could be used to compare our models.
